---
title: "hypothesize: Statistical Tests in R"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
options(digits = 3)
```

# `hypothesize`: Statistical Tests in R

`hypothesize` is a simple hypothesis testing API in R.
It is mostly designed to be used by other libraries so that they can wrap
their own hypothesis tests in a consistent way.

In this library, we define the API as a set of generics, and then provide
a simple implementation of the API for the likelihood ratio test.

## Installation

You can install the development version of `hypothesize` from GitHub with:

```{r}
# install.packages("devtools")
#devtools::install_github("queelius/hypothesize")
```

## Load the Package

```{r}
library(hypothesize)
#source("R/hypothesize.R")
```

## Core Features

`hypothesize` includes the following core functions:

- `hypothesis_test()`: Creates a hypothesis test object. You can create your
    own, since the API is defined as a set of generics. This one implements
    the API, and we also provide a simple implementation for the likelihood
    ratio test that returns this kind of object.
- `pval()`: Extracts the p-value from a hypothesis test.
- `dof()`: Retrieves the degrees of freedom associated with a hypothesis test.
- `test_stat()`: Obtains the test statistic from the hypothesis test.
- `is_significant_at()`: Determines if the hypothesis test is significant at a specified significance level.
- `lrt()`: Performs a Likelihood Ratio Test based on log-likelihood values from nested models.
- `wald_test()`: Performs a Wald test to compare a parameter estimate to a specified value.
- `z_test()`: Performs a Z-test to compare a parameter estimate to a specified value.

## Example: Using `lrt`

The `lrt` function is particularly useful for comparing nested models — where one model (the null model) is a special case of another (the alternative model).

### Scenario

Suppose we have two models that aim to explain the same dataset. Model 1 (the null model) is simpler, with fewer parameters, while Model 2 (the alternative model) includes additional parameters. We wish to test if the complexity of Model 2 is justified by a significantly better fit to the data.

### Step-by-Step Example

1. **Define Log-Likelihoods**: Assume we have calculated the log-likelihoods for both models on the same dataset. For the null model, the log-likelihood is -100, and for the alternative model, it is -99.
Assume that the difference in degrees of freedom between the two models is 2.

2. **Perform LRT**: We use `lrt` to perform the Likelihood Ratio Test.

```{r}
# Perform LRT
stat <- lrt(null_loglik = -100, alt_loglik = -96.105, dof = 3)
print(stat)
```

We show the output of the `stat` object, which includes all the relevant information about the test.
However, we might want to look at its parts independently, particularly if
we need programmatic accees to relevant parts of the test.

3. **Evaluate Significance**: Determine if the difference in log-likelihoods is significant at the 5% level.

```{r}
# Check significance
is_significant_at(stat, 0.05)
```

A negative test result indicates that the alternative model is not compatible with the data at the 5% significance level.
However, we might want to extract the test statistic, p-value, and degrees of freedom to arrive
at a more nuanced interpretation.

4. **Examine the Test Result**: Extract and examine the test statistic, p-value, and degrees of freedom to evaluate the significance.

```{r}
# Extract test statistic
test_stat(stat)

# Extract p-value
pval(stat)

# Extract degrees of freedom
dof(stat)
```

We see that the $p$-value is only slightly above our (arbitrarily) specified
$\alpha = 0.05$. This suggests that the alternative model may be reasonable
to consider, but it is not a clear-cut decision. In practice, we would
likely want to consider other factors, such as the practical significance
of the additional complexity, or collecting more data to reduce uncertainty,
before making a final decision.

## Example: Using Wald Test

The Wald test is also implemented in `hypothesize`. Tis test is used to compare
the value of a parameter to a specified value, and is often used in the context
of regression models.

```{r}
# Example: Wald Test
print(wald_test(estimate = 1.5, se = 0.5, null_value = 1))
```